---
title: "DATA399 Final Solutions"
author: "Jenna Hopkins"
date: "2026-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("C://Users/jenna/Documents/DATA399/DATA399-Final-Project/Final Project")

library(tidyverse)
library(FactoMineR)
library(factoextra)
library(umap)
library(cluster)
library(GGally)
library(plotly)
library(dbscan)
library(patchwork)
library(purrr)
library(randomForest)
library(reshape2)
library(lme4)
library(lawstat)
library(rstatix)
library(vegan)
library(nortest)
library(FSA)
library(vegan)
library(scatterplot3d)
library(mgcv)
library(tidytext)
library(pROC)
library(gt)
library(scales)
library(patchwork)
library(caret)
```

# Read in Data
First, read in the dataset and split it into train and test data (80/20 split)
```{r}
ds_fire_og = read_csv("combined_dataset.csv")

set.seed(444)
```

# Tests for Normality

```{r}
Antelope_2021 = ds_fire_og %>%
  filter(fire == "Antelope_2021")
Bald_2014 = ds_fire_og %>%
  filter(fire == "Bald_2014")
Carr_2018 = ds_fire_og %>%
  filter(fire == "Carr_2018")
McKinney_2022 = ds_fire_og %>%
  filter(fire == "McKinney_2022")
Dixie_2021 = ds_fire_og %>%
  filter(fire == "Dixie_2021")
Monument_2021 = ds_fire_og %>%
  filter(fire == "Monument_2021")
King_2014 = ds_fire_og %>%
  filter(fire == "King_2014")
NorthComplex_2020 = ds_fire_og %>%
  filter(fire == "NorthComplex_2020")
Hat_2018 = ds_fire_og %>%
  filter(fire == "Hat_2018")
Camp_2018 = ds_fire_og %>%
  filter(fire == "Camp_2018")

fire_list = list(
  Antelope_2021, Bald_2014, Carr_2018, McKinney_2022, Dixie_2021, 
  Monument_2021, King_2014, NorthComplex_2020, Hat_2018, Camp_2018
)
```

## Shapiro test
p-values indicate none of the fires are normally distributed
```{r}
shapiro_results = lapply(fire_list, function(df) shapiro.test(df$severity))
names(shapiro_results) = c("Antelope", "Bald", "Carr", "McKinney", "Dixie", 
                            "Monument", "King", "NorthComplex", "Hat", "Camp")
print(shapiro_results) 
```

## Anderson-Darling test (more sensitive to outlier data)
p-values indicate none of the fires are normally distributed
```{r}
ad_results = lapply(fire_list, function(df) ad.test(df$severity))
names(ad_results) = names(shapiro_results)
print(ad_results)
```

Since the data is not normally distributed, we will proceed by using medians instead of means


# Testing for differences in severity across fires 

## Kruskall-Wallis rank sum test 
compares median severities across fires

h0 = there is no difference of median severity between any fires
ha = there is a difference of median severity between at least two fires
```{r}
kruskal.test(severity ~ fire, data = ds_fire_og) # chi2 = 90, df = 9, p-value = 1e-15
```

The significant p-value means we reject the null hypothesis, and the large chi-squared value indicates that the 'ranks' of severity values are not distributed evenly across the fire groups


## Dunn's Test Pairwise Comparisons 
tells us which fires are different
```{r}
dunn_results = dunnTest(severity ~ fire, 
                         data = ds_fire_og, 
                         method = "holm")
summary_table = dunn_results$res[, c("Comparison", "P.adj")]

summary_table %>%
  arrange(P.adj)
summary_table %>%
  filter(P.adj <= 0.05) %>%
  nrow() # 12
```

12/45 combinations (pairs) of fires are statistically significantly different in their severity distribtions.


# Forming Fire Clusters

## Hierarchical Cluster Analysis (HCA)
is an unsupervised machine learning method that builds a hierarchy of clusters by calculating Euclidean distances, and presents the results in the form of a dendrogram. 
```{r}
fire_medians = aggregate(severity ~ fire, data = ds_fire_og, FUN = median)
dist_matrix = dist(fire_medians$severity)
clusters = hclust(dist_matrix)
plot(clusters, labels = fire_medians$fire, main = "Fire Similarity Groups")
```

We need to know the exact number of clusters and what fires should go in which cluster. Thus, the HCA plot useful going forward, but needs to be refined.

## Elbow Method
Calculate the Total Within-Cluster Sum of Squares for different numbers of clusters (k). We will test k = 1-9, then create a plot using the Elbow method with the number of clusters on the x-axis and the WSS statistic as the y-axis. This shows the number of clusters that are effectively reducing the variance in the statistic.
```{r}
row.names(fire_medians) = fire_medians$fire
vals = as.matrix(fire_medians$severity)

wss = sapply(1:9, function(k) {
  group = cutree(clusters, k = k)
  sum(sapply(unique(group), function(g) {
    sum((vals[group == g] - mean(vals[group == g]))^2)
  }))
})

plot(1:9, wss, type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares",
     main="Elbow Method for Fire Groups")
```

The Elbow method suggests that 3 clusters is the ideal number and we have reduced the statistic significantly, and having 4 or more clusters will lead to unnecessary over-fitting, where we are splitting groups that are already statistically similar. 

Now, we will extract the fires in each group and add it to the data.
```{r}
fire_groups_nums = cutree(clusters, k = 3)
fire_groupings = tibble(
  fire = fire_medians$fire,
  group = fire_groups_nums
)
fire_groupings

group_1_names = fire_groupings %>%
  filter(group == 1) %>%
  pull(fire)
group_2_names = fire_groupings %>%
  filter(group == 2) %>%
  pull(fire)
group_3_names = fire_groupings %>%
  filter(group == 3) %>%
  pull(fire)

ds_fire = ds_fire_og %>%
  mutate(
    fire_group = case_when(
      fire %in% group_1_names ~ "1",
      fire %in% group_2_names ~ "2",
      fire %in% group_3_names ~ "3"
    ),
    fire_group = factor(fire_group)
  )
```


# Unsupervised Machine Learning

First, we need to do some additional wrangling.
```{r}
# get rid of coordinates and factor all categorical variables
ds_fire = ds_fire %>%
  select(-x, -y) %>%
  mutate(
    fire = factor(fire),
    treated = factor(treated),
    fuel_model = factor(fuel_model),
    vegetation_type = factor(vegetation_type),
    fuel_model_group = factor(fuel_model_group),
    vegetation_type_group = factor(vegetation_type_group)
  ) %>%
  # get rid of severity outliers
  group_by(fire) %>%
  filter(
    severity > quantile(severity, 0.01, na.rm=TRUE),
    severity < quantile(severity, 0.99, na.rm=TRUE)
  ) %>%
  ungroup() %>%
  # classify severity
  mutate(
    severity_class = case_when(
      severity <= -100 ~ "Increased_Vegetation",
      severity >= -100 & severity <= 100 ~ "Low_Severity",
      severity >= 100 & severity <= 270 ~ "Moderate_Severity",
      TRUE ~ "High Severity"
    ),
    severity_class = factor(severity_class)
  ) %>%
  # get rid of missing data
  filter(!is.na(severity_class), 
         !is.na(fire_group),
         !is.na(vpdmax),
         !is.na(elevation_m),
         !is.na(fuel_model_group),
         !is.na(vegetation_type_group)
  )
```


### Non-Metric Multidimensional Scaling (NMDS) - Non-linear:
Non-linear dimensionality reduction technique used to visualize similarity or dissimilarity between complex datasets. Focuses on preserving the rank-order of distances between samples, making it ideal for ecological and high-dimensional data that do not follow linear distributions. Used as a tool for visualization and EDA rather than direct prediction, as it helps identify clusters, patterns, and gradients in data.

The Bray-Curtis distance, which is used to calculate the distance for the NMDS method, requires a numerical matrix. Since the Gower's distance method effectively handles both numerical and categorical data and creates distance between points, it is used on our data prior to feeding the data into the NMDS function.

```{r}
# selecting variables?
nmds_data = ds_fire %>%
  select(
    #ppt,
    #tmax,
    vpdmax,
    #canopy_cover,
    #slope_deg,
    #aspect_deg,
    elevation_m,
    fuel_model_group,
    vegetation_type_group) %>%
  mutate(across(where(is.character), as.factor))

# using Gower's distance method
dist_matrix = daisy(nmds_data, metric = "gower")

# running nmds
nmds_result_3d = metaMDS(dist_matrix, k = 3, trymax = 25, autotransform = FALSE)
nmds_matrix_3d = as.matrix(nmds_result_3d$points)

fit = envfit(nmds_matrix_3d, 
             fire[, c("severity_class", "fire_group", "vpdmax", "elevation_m", "fuel_model_group", "vegetation_type_group")], 
              perm = 999)

print(fit)
```


Filter Vectors for significance (numerical)
```{r}
sig_vec_idx <- which(fit$vectors$pvals < 0.05)
vec_coords_sig <- as.data.frame(fit$vectors$arrows[sig_vec_idx, , drop = FALSE])
vec_coords_sig$variable <- rownames(vec_coords_sig)
```


Filter Factors for significance(categorical)
```{r}
sig_fact_idx <- which(fit$factors$pvals < 0.05)
# This identifies which CATEGORIES (e.g., 'Timber') belong to significant groups
fact_coords_sig <- as.data.frame(fit$factors$centroids[sig_fact_idx, , drop = FALSE])
fact_coords_sig$variable <- rownames(fact_coords_sig)
```


visualize nmds code:


## Non-Linear Clustering
UMAPS, Gower's Distances used if it is found that linear relationships do not accurately represent the data and the relationship between the variables. 

First, more data wrangling and EDA
```{r}
unsup_ds = ds_fire %>%
  mutate(across(where(is.numeric), ~ifelse(is.infinite(.), NA, .)))

num_vars = unsup_ds %>%
  select(severity, ppt, tmax, vpdmax, canopy_cover, slope_deg, aspect_deg, elevation_m)
cor_matrix = cor(num_vars, use="pairwise.complete.obs")
corrplot::corrplot(cor_matrix, method="color")
GGally::ggpairs(num_vars)
severity_cor = cor_matrix["severity", ]
severity_cor

fire_list = unsup_ds %>%
  group_split(fire)
```


### Gower's Distance, UMAPS, HBDSCAN

Function to run all 3:
```{r}
run_fire_pipeline = function(data){
  fire_name = unique(data$fire)
  feature_cols = c(
    #"ppt",
    "tmax",
    #"vpdmax",
    "canopy_cover",
    #"slope_deg",
    #"aspect_deg",
    "elevation_m",
    "fuel_model_group",
    "vegetation_type_group"
  )
  features = data %>%
    select(all_of(feature_cols)) %>%
    drop_na()
  n = nrow(features)
  
  # Skip fires that are too small
  if(n < 10){
    return(NULL)
  }
  
  features = features %>%
  dplyr::mutate(
    dplyr::across(where(is.character), as.factor)
  )

  # GOWER DISTANCE
  gower_dist = cluster::daisy(features, metric = "gower")
 
  # UMAP (dynamic neighbors)
  neighbors = min(10, n - 1)
  umap_config = umap::umap.defaults
  umap_config$n_neighbors = neighbors
  umap_res = umap::umap(as.matrix(gower_dist), config = umap_config)
  coords = as.data.frame(umap_res$layout)
  colnames(coords) = c("UMAP1","UMAP2")

  # HDBSCAN
  minpts = max(5, floor(n * 0.05))
  hdb = dbscan::hdbscan(coords, minPts = minpts)
  coords$cluster = factor(hdb$cluster)
  coords$fire = fire_name
  cluster_data = data %>%
    slice(1:n) %>%
    mutate(cluster = coords$cluster)
  list(
    coords = coords,
    cluster_data = cluster_data
  )
}
```


Calling the function:
```{r}
results = purrr::map(fire_list, run_fire_pipeline)
results = results[!sapply(results, is.null)]

umap_results = purrr::map_dfr(results, "coords")
cluster_data = purrr::map_dfr(results, "cluster_data")
```


Visualizing the results
```{r}
ggplot(umap_results,
       aes(UMAP1, UMAP2, color = cluster)) +
  geom_point(alpha = .7) +
  facet_wrap(~fire) +
  theme_minimal() +
  labs(title = "UMAP Clusters by Fire")

ggplot(cluster_data,
       aes(cluster, severity, fill = cluster)) +
  geom_boxplot() +
  facet_wrap(~fire) +
  theme_minimal() +
  labs(title = "Severity Distribution by Cluster")

cluster_summary = cluster_data %>%
  group_by(fire, cluster) %>%
  summarise(
    across(
      c(tmax, canopy_cover,
        elevation_m),
      median,
      na.rm = TRUE
    ),
    .groups = "drop"
  )
cluster_summary

cluster_summary %>%
  pivot_longer(
    -c(fire, cluster),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(cluster, variable, fill = value)) +
    geom_tile() +
    facet_wrap(~fire, scales = "free") +
    scale_fill_viridis_c() +
    theme_minimal()

ggplot(cluster_data,
       aes(cluster, fill = fuel_model_group)) +
  geom_bar(position = "fill") +
  facet_wrap(~fire) +
  theme_minimal() +
  labs(
    y = "Proportion",
    title = "Fuel Model Composition by Cluster"
  )
```


#### Characteristics of each cluster

```{r}
cluster_classifications = cluster_data %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    elevation_m = median(elevation_m, na.rm = TRUE),
    vpdmax = median(vpdmax, na.rm = TRUE),
    canopy_cover = median(canopy_cover, na.rm = TRUE),
    slope_deg = median(slope_deg, na.rm = TRUE),

    # proportions for categorical
    fuel_grass = mean(fuel_model_group == "Grass", na.rm = TRUE),
    fuel_shrub = mean(fuel_model_group == "Shrub", na.rm = TRUE),
    fuel_timberunderstory = mean(fuel_model_group == "Timber-Understory", na.rm = TRUE),
    fuel_timberlitter = mean(fuel_model_group == "Timber Litter", na.rm = TRUE),
    fuel_grassshrub = mean(fuel_model_group == "Grass-Shrub", na.rm = TRUE),
    fuel_nonburn = mean(fuel_model_group == "Non-Burnable", na.rm = TRUE),

    veg_tree = mean(vegetation_type_group == "Tree", na.rm = TRUE),
    veg_shrub = mean(vegetation_type_group == "Shrub", na.rm = TRUE),
    veg_developed = mean(vegetation_type_group == "Developed", na.rm = TRUE),
    veg_herb = mean(vegetation_type_group == "Herb", na.rm = TRUE),
    veg_sparse = mean(vegetation_type_group == "Sparse", na.rm = TRUE),
    veg_water = mean(vegetation_type_group == "Water", na.rm = TRUE),
    veg_ag = mean(vegetation_type_group == "Agriculture", na.rm = TRUE),
    veg_barren = mean(vegetation_type_group == "Barren", na.rm = TRUE),

    severity = median(severity, na.rm = TRUE),

    .groups = "drop"
  )

cluster_classifications
```


Visualizing the clusters:
```{r}
# turn it into a table then dataframe to make it easier to work with
cluster_t = as.data.frame(t(cluster_classifications), stringsAsFactors = FALSE)

# make the rownames under one 'variable' column
cluster_t = tibble::rownames_to_column(cluster_t, var = "variable")
# take the cluster section of variable out, it's already implied
cluster_t = cluster_t[cluster_t$variable != "cluster", ]
# take the V out of the cluster column names
colnames(cluster_t) = gsub("^V", "", colnames(cluster_t))

# arrange by severity
cluster_t = cluster_t %>%
  arrange(desc(variable == "severity"))

# turn variables into numberic and round to hundredth
cluster_t[,-1] = lapply(cluster_t[,-1], as.numeric)
cluster_t[,-1] = lapply(cluster_t[,-1], function(x) round(as.numeric(x), 2))

# rename all of the 'variable' rows so the plot is easier to read
cluster_t = cluster_t %>%
  mutate(variable = recode(variable,
    severity = "Severity",
    n = "Number of Points",
    vpdmax = "VPD",
    elevation_m = "Elevation",
    canopy_cover = "Canopy Cover",
    slope_deg = "Slope",
    aspect_deg = "Aspect",
    fuel_grass = "Fuel: Grass",
    fuel_shrub = "Fuel: Shrub",
    fuel_timberunderstory = "Fuel: Timer-Understory",
    fuel_timberlitter = "Fuel: Timber-Litter",
    fuel_grassshrub = "Fuel: Grass-Shrub",
    fuel_nonburn = "Fuel: Non-Burnable",
    veg_shrub = "Veg: Shrub",
    veg_tree = "Veg: Tree",
    veg_developed = "Veg: Developed",
    veg_herb = "Veg: Herb",
    veg_sparse = "Veg: Sparse",
    veg_water = "Veg: Water",
    veg_ag = "Veg: Agriculture",
    veg_barren = "Veg: Barren"
  ))

# pivot longer
df = cluster_t
#df[,-1] <- lapply(df[,-1], as.numeric)
df_long = df %>%
  pivot_longer(
    cols = -variable,
    names_to = "cluster",
    values_to = "value"
  )

# scale the values of each row relative to each other for the coloring in the plot
df_long = df_long %>%
  group_by(variable) %>%
  mutate(
    scaled = (value - min(value, na.rm = TRUE)) /
             (max(value, na.rm = TRUE) - min(value, na.rm = TRUE))
  ) %>%
  ungroup()


# set the order that variable and cluster should appear in
df_long$variable = factor(
  df_long$variable,
  levels = c(
    "Severity",
    "Number of Points",
    "VPD",
    "Elevation",
    "Slope",
    "Aspect",
    "Canopy Cover",
    "Fuel: Grass",
    "Fuel: Shrub",
    "Fuel: Timer-Understory",
    "Fuel: Timber-Litter",
    "Fuel: Grass-Shrub",
    "Fuel: Non-Burnable",
    "Veg: Shrub",
    "Veg: Tree",
    "Veg: Developed",
    "Veg: Herb",
    "Veg: Sparse",
    "Veg: Water",
    "Veg: Agriculture",
    "Veg: Barren"
  )
)
df_long$variable <- factor(df_long$variable, levels = rev(levels(df_long$variable)))
df_long$cluster = factor(
  df_long$cluster,
  levels = c(
    "1",
    "2",
    "3",
    "4",
    "5",
    "6",
    "7",
    "8",
    "9",
    "10",
    "11"
  )
)

# final visualization of the heatmap and top part of the figure
heatmap_plot = ggplot(df_long, aes(x = cluster, y = variable, fill = scaled)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "firebrick1") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  theme_bw() +
  labs(
    title = "Fire Regime Cluster Environmental Profiles and Recommended Treatments",
    x = "Cluster",
    y = "Variable",
    fill = "Relative\nIntensity"
  ) +
  guides(fill = guide_colorbar(
    title.position = "left",
    title.hjust = 0.5,
    barwidth = 10,
    barheight = 0.5
    )) +
  theme(
    legend.position = "top",
    legend.box = "horizontal",
    legend.key.height = unit(0.1, "cm"),
    legend.key.width = unit(1, "cm"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8, margin = ggplot2::margin(r = 10)),
    plot.margin = ggplot2::margin(1, 1, 1, 1),
    plot.title = element_text(hjust = 2)
  )



# bottom part of the figure, treatments for each cluster
treatments = data.frame(
  cluster = factor(1:11),
  treatment = c(
    "Rx Fire + Grazing",
    "Rx Fire + Fuel Breaks",
    "Thin + Rx Fire",
    "Rx Fire + Patch Thin",
    "Rx Fire + Fuel Breaks",
    "Thin + Rx Fire",
    "Thin + Rx Fire",
    "Defensible Space + Fuel Breaks",
    "Intensive Thin + Fuel Breaks",
    "Thin + Managed Fire",
    "Edge Thin + Defensible Space"
  )
)

treatments$cluster = factor(treatments$cluster, levels = levels(df_long$cluster))
treatments$treatment = str_wrap(treatments$treatment, width = 7)

# making the bottom plot with all the treatments
p_treat = ggplot(treatments, aes(x = cluster, y = 1, label = treatment)) +
  geom_vline(xintercept = seq(1.5, 10.5, by = 1), color = "grey80", linewidth = 0.4) +
  geom_text(
    aes(label = treatment),
    size = 2.3,
    hjust = 0,
    nudge_x = -0.35
  ) +
  theme_void() +
  theme(
    plot.margin = ggplot2::margin(1, 1, 1, 1),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 9
    )
  ) +
  labs(subtitle = "Recommended Treatments")


# putting both plots together
(heatmap_plot / p_treat) +
  plot_layout(heights = c(4, 1)) 
```


```{r}
clusters_long = cluster_classifications %>%
  pivot_longer(
    cols = starts_with("fuel_") | starts_with("veg_"),
    names_to = c("fuel_veg", "category"),
    names_sep = "_",
    values_to = "prop"
  )

clusters_long %>%
  filter(cluster == 0) %>%
  ggplot(aes(x = reorder_within(category, -prop, fuel_veg), y = prop)) +
  geom_col() +
  facet_wrap(~fuel_veg, scales = "free_x") +
  scale_x_reordered() +
  theme(axis.text.x = element_text(angle = 90))
  
clusters_long %>%
  filter(fuel_veg == "fuel") %>%
  ggplot(aes(reorder_within(category, -prop, fuel_veg), y = prop, fill = severity)) +
  geom_col() +
  facet_wrap(~cluster) +
  scale_x_reordered() +
  theme(axis.text.x = element_text(angle = 90))

clusters_long %>%
  filter(fuel_veg == "veg") %>%
  ggplot(aes(reorder_within(category, -prop, fuel_veg), y = prop, fill = severity)) +
  geom_col() +
  facet_wrap(~cluster) +
  scale_x_reordered() +
  theme(axis.text.x = element_text(angle = 90))
```



- Cluster 0 (33.5 severity): 1292 elevation, 37 vpd, 45 canopy cover, 12 slope, majority timber fuel, majority tree vegetation
- Cluster 1 (-2 severity): 1379 elevation, 39 vpd, 0 canopy cover, 7 slope, majority grass shrub fuel, majority shrub tree vegetation
- Cluster 2 (28.5 severity): 1514 elevation, 36 vpd, 35 canopy cover, 10 slope, majority grass-shrub and timber-litter fuel, majority tree vegetation
- Cluster 3 (56 severity): 1611 elevation, 33 vpd, 45 canopy cover, 10 slope, majority timber fuel, majority tree vegetation
- Cluster 4 (36 severity): 1042 elevation, 42 vpd, 45 canopy cover, 14 slope, majority timber fuel, major majority tree vegetation
- Cluster 5 (31 severity): 967 elevation, 41 vpd, 55 canopy cover, 17 slope, major majority timber fuel, major majority tree vegetation
- Cluster 6 (8 severity): 1024 elevation, 41 vpd, 35 canopy cover, 15 slope, major majority timber understory fuel, major majority tree vegetation
- Cluster 7 (26.5 severity): 1300 elevation, 42 vpd, 35 canopy cover, 15 slope, majority timber understory and grass shrub fuel, majority tree vegetation
- Cluster 8 (27 severity): 1667 elevation, 34 vpd, 45 canopy cover, 19 slope, major majority timber fuel, all tree vegetation
- Cluster 9 (16.5 severity): 1351 elevation, 35 vpd, 45 canopy cover, 12 slope, all timber-litter fuel, all tree vegetation
- Cluster 10 (60 severity): 1715 elevation, 39 vpd, 45 canopy cover, 8 slope, all timber-litter fuel, all tree vegetation
- Cluster 11 (-6.5 severity): 1499 elevation, 43 vpd, 40 canopy cover, 3 slope, all timber-litter fuel, almost all tree vegetation


Cluster analysis revealed that burn severity is structured by distinct environmental–fuel regimes. Higher severity clusters were primarily associated with timber-dominated fuel complexes (including timber litter and understory fuels), moderate canopy cover, and higher elevation forest environments. In contrast, lower severity clusters tended to occur in more open or mixed fuel environments, including grass–shrub and low canopy conditions. These results suggest that fuel continuity and forest structure are more strongly associated with severity patterns than topographic variables such as slope.



Testing the results:
```{r}
aov(severity ~ cluster, data = cluster_data)
chisq.test(table(cluster_data$fuel_model_group,
                 cluster_data$cluster))
```



# Solution Models

## Solution 1: Predicting for severity

```{r}
# 80/20 train test split
train_idx = sample(1:nrow(ds_fire), 0.8 * nrow(ds_fire))
train_raw = ds_fire[train_idx, ]
test_raw = ds_fire[-train_idx, ]

# making severity binary class from the 4-class variable
test_raw = test_raw %>%
  mutate(
    severity_binary = case_when(
      severity_class == "High Severity" ~ 1,
      severity_class == "Moderate_Severity" ~ 1,
      severity_class == "Low_Severity" ~ 0,
      severity_class == "Increased_Vegetation" ~ 0
    ),
    severity_binary = factor(severity_binary)
  )
train_raw = train_raw %>%
  mutate(
    severity_binary = case_when(
      severity_class == "High Severity" ~ 1,
      severity_class == "Moderate_Severity" ~ 1,
      severity_class == "Low_Severity" ~ 0,
      severity_class == "Increased_Vegetation" ~ 0
    ),
    severity_binary = factor(severity_binary)
  )

# fitting a random forest model
rf_bin = randomForest(
  severity_binary ~ ppt + tmax + vpdmax + elevation_m,
  data = train_raw,
  ntree = 1000,        # more trees = more stable
  mtry = 3,            
  nodesize = 5,
  importance = TRUE
)

# using the test data to determine how good it is
pred = predict(rf_bin, test_raw, type = "prob")[,2]
pred_class = ifelse(pred > 0.5, 1, 0)

mean(pred_class == test_raw$severity_binary)
roc(test_raw$severity_binary, pred)
varImpPlot(rf_bin)
plot(roc(test_raw$severity_binary, pred), auc.polygon = TRUE)
```

AUC measures a model's ability to distinguish between two classes. It represents the probability that a randomly chosen positive instance will be ranked higher by the model than a randomly chosen negative instance. The plot represents the trade-off between Sensitivity (True Positive Rate) and Specificity (1 - False Positive Rate), a curve "hugging" the top-left corner indicates a high AUC


### Generalize to new fires

Data wrangling that happened to the original 10 fire dataset
```{r}
new_fires_og = read_csv("new_fires_dataset.csv")

new_fires = new_fires_og %>%
  mutate(
    fire = factor(fire),
    treated = factor(treated),
    fuel_model = factor(fuel_model),
    vegetation_type = factor(vegetation_type),
    fuel_model_group = factor(fuel_model_group),
    vegetation_type_group = factor(vegetation_type_group)
  ) %>%
  # get rid of severity outliers
  group_by(fire) %>%
  filter(
    severity > quantile(severity, 0.01, na.rm=TRUE),
    severity < quantile(severity, 0.99, na.rm=TRUE)
  ) %>%
  ungroup() %>%
  # classify severity
  mutate(
    severity_class = case_when(
      severity <= -100 ~ "Increased_Vegetation",
      severity >= -100 & severity <= 100 ~ "Low_Severity",
      severity >= 100 & severity <= 270 ~ "Moderate_Severity",
      TRUE ~ "High Severity"
    ),
    severity_class = factor(severity_class)
  ) %>%
  # get rid of missing data
  filter(!is.na(severity_class), 
         !is.na(vpdmax),
         !is.na(elevation_m),
         !is.na(fuel_model_group),
         !is.na(vegetation_type_group)
  )

new_fires = new_fires %>%
  mutate(
    severity_binary = case_when(
      severity_class == "High Severity" ~ 1,
      severity_class == "Moderate_Severity" ~ 1,
      severity_class == "Low_Severity" ~ 0,
      severity_class == "Increased_Vegetation" ~ 0
    ),
    severity_binary = factor(severity_binary)
  )
```

Making predictions with the new fires
```{r}
pred_new_fires = predict(rf_bin, new_fires, type = "prob")[,2]
pred_class_new_fires = ifelse(pred_new_fires > 0.5, 1, 0)

new_fires$pred_prob = pred_new_fires
new_fires$pred_class = pred_class_new_fires

# calculating some stats to use in subsequent plots
new_fires = new_fires %>%
  mutate(
    mismatch = ifelse(severity_binary == 1 & treated == 0, 1, 0),
    case = case_when(
      severity_binary == 1 & treated == 1 ~ "High severity + treated",
      severity_binary == 1 & treated == 0 ~ "High severity + not treated",
      severity_binary == 0 & treated == 1 ~ "Low severity + treated",
      TRUE ~ "Low severity + not treated"
    ),
    correct = ifelse(pred_class == severity_binary, 1, 0),
    model_helpful = ifelse(pred_class == 1 & severity_binary == 1 & treated == 0,1, 0),
    model_worse = ifelse(pred_class == 0 & severity_binary == 1 & treated == 1, 1, 0),
    case_type = case_when(
      model_worse == 1 ~ "Model worse",
      model_helpful == 1 ~ "Model helpful",
      TRUE ~ "Other"
    )
  )
```


Visualizing results with the new fires

Severity risk vs treatment locations
```{r}
McCash_subset = new_fires %>%
  filter(fire == "McCash_2021") 
Park_subset = new_fires %>%
  filter(fire == "Park_2024")

ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "McCash Fire (2021)")


ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  geom_point(
    data = subset(McCash_subset, treated == 1),
    color = "cyan",
    size = 0.7,
    alpha = 0.7
  ) +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "Park Fire (2024)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  geom_point(
    data = subset(Park_subset, treated == 1),
    color = "cyan",
    size = 0.7,
    alpha = 0.7
  ) +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "Park Fire (2024)")
```


Predicted class
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(pred_class)), size = 0.5) +
  theme_minimal() +
  labs(color = "Predicted Class",
       subtitle = "McCash Fire (2021)")



ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(pred_class)), size = 0.5) +
  theme_minimal() +
  labs(color = "Predicted Class",
       subtitle = "Park Fire (2024)")
```


High-risk areas without treatment
```{r}

ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(mismatch)), size = 0.5) +
  scale_color_manual(values = c("0" = "grey80", "1" = "red"),
                     labels = c("OK", "High Risk Untreated")) +
  theme_minimal() +
  labs(title = "High-Risk Areas Without Treatment",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(mismatch)), size = 0.5) +
  scale_color_manual(values = c("0" = "grey80", "1" = "red"),
                     labels = c("OK", "High Risk Untreated")) +
  theme_minimal() +
  labs(title = "High-Risk Areas Without Treatment",
       subtitle = "Park Fire (2024)")
```


Actual wildfire severity with treatment locations
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(severity_binary)), size = 0.5) +
  geom_point(
    data = subset(McCash_subset, treated == 1),
    color = "cyan",
    size = 0.6,
    alpha = 0.6
  ) +
  scale_color_manual(values = c("grey80", "red"),
                     labels = c("Low severity", "High severity")) +
  theme_minimal() +
  labs(title = "Actual Wildfire Severity with Treatment Locations",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(severity_binary)), size = 0.5) +
  geom_point(
    data = subset(Park_subset, treated == 1),
    color = "cyan",
    size = 0.6,
    alpha = 0.6
  ) +
  scale_color_manual(values = c("grey80", "red"),
                     labels = c("Low severity", "High severity")) +
  theme_minimal() +
  labs(title = "Actual Wildfire Severity with Treatment Locations",
       subtitle = "Park Fire (2024)")
```


model prediction accuracy across landscape
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(correct)), size = 0.5) +
  scale_color_manual(values = c("red", "green"),
                     labels = c("Incorrect", "Correct")) +
  theme_minimal() +
  labs(title = "Model Prediction Accuracy Across Landscape",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(correct)), size = 0.5) +
  scale_color_manual(values = c("red", "green"),
                     labels = c("Incorrect", "Correct")) +
  theme_minimal() +
  labs(title = "Model Prediction Accuracy Across Landscape",
       subtitle = "Park Fire (2024)")
```


model effectiveness
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = case_type), size = 0.5) +
  scale_color_manual(
    values = c(
      "Model helpful" = "green",
      "Model worse" = "red",
      "Other" = "grey85"
    )
  ) +
  theme_minimal() +
  labs(
    title = "Model Effectiveness for the McCash Fire (2021)",
    subtitle = "Red = high severity + treated areas missed by model",
    color = "Model Outcome"
  )



ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = case_type), size = 0.5) +
  scale_color_manual(
    values = c(
      "Model helpful" = "green",
      "Model worse" = "red",
      "Other" = "grey85"
    )
  ) +
  theme_minimal() +
  labs(
    title = "Model Effectiveness for the Park Fire (2024)",
    subtitle = "Red = high severity + treated areas missed by model",
    color = "Model Outcome"
  )
```


determining the efficiency of treatment for model and actual
```{r}

compute_treatment_efficiency <- function(df, fire_name) {

  n_treated <- sum(df$treated == 1)

  real_hit_rate <- df %>%
    filter(treated == 1) %>%
    summarise(hit = mean(severity_binary == 1)) %>%
    pull(hit)

  model_treated <- df %>%
    arrange(desc(pred_prob)) %>%
    slice(1:n_treated)

  model_hit_rate <- mean(model_treated$severity_binary == 1)

  data.frame(
    fire = fire_name,
    method = c("Actual treated area", "Model-based treated area"),
    hits = c(real_hit_rate, model_hit_rate)
  )
}

results <- bind_rows(
  compute_treatment_efficiency(McCash_subset, "McCash (2021)"),
  compute_treatment_efficiency(Park_subset, "Park (2024)")
)


ggplot(results, aes(x = method, y = hits, fill = method)) +
  geom_bar(stat = "identity") +
  facet_wrap(~fire) +
  theme_bw() +
  labs(
    title = "Efficiency of Treatment Placement Across Fires",
    y = "High-severity pixels captured",
    x = NULL
  ) +
  scale_y_continuous(labels=scales::percent_format(), limits = c(0, 1)) +
  theme(legend.position = "none")
```




## Solution 2: Clustering/Classification Model

post-hoc classification of unsupervised structure, testing whether clusters correspond to predictable environmental regimes - Can we reconstruct these groupings from raw variables?

testing stability and separability of clusters in feature space
```{r}
cluster_data = na.omit(cluster_data)
train_idx = sample(1:nrow(cluster_data), 0.8 * nrow(cluster_data))
train = cluster_data[train_idx, ]
test  = cluster_data[-train_idx, ]

predictors = c(
  "elevation_m",
  "canopy_cover",
  "slope_deg",
  "vpdmax",
  "tmax",
  "fuel_model_group",
  "vegetation_type_group",
  "ppt",
  "aspect_deg"
)
```


```{r}
model1 = randomForest(cluster ~ ., data = train[, c(predictors, "cluster")])
pred1 = predict(model1, test)
mean(pred1 == test$cluster)
cm1 = table(Predicted = pred1, Actual = test$cluster)
cm1
diag(cm1) / colSums(cm1)

varImpPlot(model1)
```



```{r}
model2 = randomForest(
  cluster ~ .,
  data = train[, c(predictors, "cluster")],
  strata = train$cluster,
  sampsize = rep(min(table(train$cluster)), length(unique(train$cluster)))
)
pred2 = predict(model2, test)
mean(pred2 == test$cluster)

cm2 = table(Predicted = pred2, Actual = test$cluster)
cm2
diag(cm2) / colSums(cm2)
```


```{r}
train_control = trainControl(method = "cv", number = 5)

model3 = train(
  cluster ~ .,
  data = train[, c(predictors, "cluster")],
  method = "rf",
  trControl = train_control
)

pred3 = predict(model3, test)
mean(pred3 == test$cluster)
cm3 = table(Predicted = pred3, Actual = test$cluster)
cm3
diag(cm3) / colSums(cm3)

plot(varImp(model3))
```


Cross-validated classification accuracy was approximately 67%, indicating moderate but meaningful separability of unsupervised fire regime clusters based on environmental predictors.



# Recommendations figure

```{r}
# setting up structure and text of the flowchart
nodes = data.frame(
  x = c(1,1,1,1,1,1,1),
  y = c(7,6,5,4,3,2,1),
  label = c(
    "Severity, Treatment, and Environment Data",
    "Random Forest Risk Model (Solution 1)",
    "High-Risk Areas Identified",
    "Regime Classification Model (Solution 2)",
    "Regime-Specific Treatment Selection",
    "Targeted Treatment Implementation",
    "Reduced High-Severity Fire Outcomes"
  )
)

# making the last one bold
nodes$label[7] = expression(bold("Reduced High-Severity Fire Outcomes"))

# plotting it as a segment graph
ggplot(nodes, aes(x = x, y = y)) +
  geom_text(aes(label = label), size = 4, lineheight = 0.9) +
  geom_segment(aes(x = 1, xend = 1, y = 6.8, yend = 6.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 5.8, yend = 5.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 4.8, yend = 4.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 3.8, yend = 3.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 2.8, yend = 2.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 1.8, yend = 1.2), arrow = arrow(length = unit(0.2, "cm"))) +
  theme_void() +
  labs(
    title = expression(bold("Integrated Decision Framework for Wildfire Treatment Allocation and Design"))
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```




