---
title: "Wildfire Analysis using Machine Learning and Data Analysis"
author: "Jenna Hopkins, Jerrick Little, Anujan Tennathur"
date: "2026-04-28"
output: pdf_document
---


#' Reading in Dataset
```{r}
#setwd("C://Users/anujantennathur/Documents/GitHub/DATA399-Final-Project")
#setwd("C://Users/jenna/Documents/DATA399/DATA399-Final-Project")
setwd("C://Users/jerri/OneDrive/Documents/DATA399/DATA399-Final-Project/Final Project")

set.seed(444)

# Libraries

library(readr)
library(cluster)
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(umap)
library(GGally)
library(plotly)
library(dbscan)
library(patchwork)
library(purrr)
library(randomForest)
library(reshape2)
library(lme4)
library(lawstat)
library(rstatix)
library(vegan)
library(nortest)
library(FSA)
library(vegan)
library(scatterplot3d)
library(ggplot2)
library(sf)
library(dplyr)
library(viridis)
library(maps)
library(ggrepel)
library(mgcv)
library(tidytext)
library(pROC)
library(gt)
library(scales)
library(patchwork)
library(caret)

ds_fire = read_csv("combined_dataset.csv")
```


# Base Visualizations, EDA

## Histograms
```{r}
ds_fire_og %>%
  ggplot(aes(x = severity)) +
  geom_histogram(aes(y = after_stat(density)), 
                 binwidth = 25,
                 color = "cyan",
                 fill = "black") +
  # geom_density(color = "red", size = 1) +
  coord_cartesian(xlim = c(-1000, 1300)) +
  labs(
    title = "Distribution of Severity -- Excludes Outliers",
    x = "Severity (Relative Differenced Normalized Burned Ratio)",
    y = ""
  ) +
  facet_wrap(~fire) +
  theme_minimal()


ds_fire_og %>%
  ggplot(aes(x = canopy_cover)) +
  geom_bar()+
  # geom_density(color = "red", size = 1) +
  coord_cartesian(xlim = c(15, 60)) +
  labs(
    title = "Distribution of Canopy Cover",
    x = "Canopy Cover (in Percentage)",
    y = ""
  ) + 
  facet_wrap(~fire) + 
  theme_minimal()
```

## Severity levels of each fire within California
```{r}
ca = map_data("state") %>% filter(region == "california")
# Create one label point per fire at the centroid of its points
fire_labels = ds_fire_og %>%
  group_by(fire) %>%
  summarise(
    x = mean(x, na.rm = TRUE),
    y = mean(y, na.rm = TRUE)
  ) %>%
  mutate(fire_label = gsub("_", " ", fire))

ds_fire_map = ds_fire_og %>%
  mutate(
    severity_class = case_when(
      severity <= -100 ~ "Increased_Vegetation",
      severity >= -100 & severity <= 100 ~ "Low_Severity",
      severity >= 100 & severity <= 270 ~ "Moderate_Severity",
      TRUE ~ "High Severity"
    ),
    severity_class = factor(severity_class)
  )

ggplot() +
  geom_polygon(data = ca, aes(x = long, y = lat, group = group),
               fill = "grey90", color = "grey50") +
  geom_point(data = ds_fire_map, aes(x = x, y = y, color = severity_class),
             size = 0.8, alpha = 0.7) +
  geom_label_repel(data = fire_labels, aes(x = x, y = y, label = fire_label),
                   size = 3,
                   fontface = "bold",
                   fill = "white",
                   color = "black",
                   label.padding = unit(0.15, "lines"),
                   box.padding = unit(0.4, "lines"),
                   max.overlaps = 20) +
  scale_color_manual(values = c(
    "Increased_Vegetation" = "#1a9641",
    "Low_Severity"         = "#a1dab4",
    "Moderate_Severity"    = "#fd8d3c",
    "High Severity"        = "#bd0026"
  ), labels = c("Increased_Vegetation" = "Increased Vegetation",
                "Low_Severity" = "Low Severity",
                "Moderate_Severity" = "Moderate Severity",
                "High Severity" = "High Severity")) +
  coord_fixed(1.3, xlim = c(-125, -119), ylim = c(38, 42)) +
  theme_minimal() +
  labs(title = "Burn Severity Across 10 Northern California Fires",
       color = "Severity Class", x = "Longitude", y = "Latitude") +
  theme(legend.position = "top", legend.key.size = unit(5, "mm"))+
  guides(
    color = guide_legend(
      override.aes = list(size = 5),
      title.theme = element_text(face = "bold", size = 11),
      label.theme = element_text(size = 10)
    ))
```

# Tests for Normality on Each Fire

Creating a list of points from each fire so each can be individually tested
```{r}
Antelope_2021 = ds_fire_og %>%
  filter(fire == "Antelope_2021")
Bald_2014 = ds_fire_og %>%
  filter(fire == "Bald_2014")
Carr_2018 = ds_fire_og %>%
  filter(fire == "Carr_2018")
McKinney_2022 = ds_fire_og %>%
  filter(fire == "McKinney_2022")
Dixie_2021 = ds_fire_og %>%
  filter(fire == "Dixie_2021")
Monument_2021 = ds_fire_og %>%
  filter(fire == "Monument_2021")
King_2014 = ds_fire_og %>%
  filter(fire == "King_2014")
NorthComplex_2020 = ds_fire_og %>%
  filter(fire == "NorthComplex_2020")
Hat_2018 = ds_fire_og %>%
  filter(fire == "Hat_2018")
Camp_2018 = ds_fire_og %>%
  filter(fire == "Camp_2018")

fire_list = list(
  Antelope_2021, Bald_2014, Carr_2018, McKinney_2022, Dixie_2021, 
  Monument_2021, King_2014, NorthComplex_2020, Hat_2018, Camp_2018
)
```


## Shapiro test and Anderson Darling test
Note: Anderson Darling test is even more sensitive to outlier data compared to Shapiro test.
```{r}
shapiro_results = lapply(fire_list, function(df) shapiro.test(df$severity))
names(shapiro_results) = c("Antelope", "Bald", "Carr", "McKinney", "Dixie", 
                            "Monument", "King", "NorthComplex", "Hat", "Camp")
print(shapiro_results) 


ad_results = lapply(fire_list, function(df) ad.test(df$severity))
names(ad_results) = names(shapiro_results)
print(ad_results)

```

Both tests indicate the data is not normally distributed (low p-values and A statistic). We will proceed with a Kruskal-Wallis test instead to compare medians across fire groups.


# Testing for differences in severity across fires 

## Kruskal-Wallis rank sum test 
Compares median severities across fires.

h0 = there is no difference of median severity between any fires

ha = there is a difference of median severity between at least two fires
```{r}
kruskal.test(severity ~ fire, data = ds_fire_og)
```

Kruskal-Wallis chi-squared = 90.82, df = 9, p-value = 1.115e-15

The result indicates that there is a statistically significant difference in severity in AT LEAST two of the fires. 

- significant p value (<0.05), reject null hypothesis, meaning that median severity is different in at least one fire compared to others. 
- large chi-squared value (90.82). Indicates that the "ranks" of severity values are not distributed evenly across the fire groups.

The Kruskal-Wallis test provides a result that proves differences exist, but cannot provide which fires are differing from each other or not. This is the reason for Dunn's test, which will do pairwise comparisons.

## Dunn's Test Pairwise Comparisons
Tells us which fires are different
```{r}
dunn_results = dunnTest(severity ~ fire, 
                         data = ds_fire_og, 
                         method = "holm")
print(dunn_results)

summary_table = dunn_results$res[, c("Comparison", "P.adj")]
summary_table
```

Statistically significantly different fires:
```{r}
significant_fires = subset(dunn_results$res, P.adj < 0.05)
significant_fires = significant_fires[order(significant_fires$P.adj), ]
print(significant_fires[, c("Comparison", "P.adj")])
```

Statistically insignificantly different fires:
```{r}
similar_fires = subset(dunn_results$res, P.adj > 0.05)
similar_fires = similar_fires[order(similar_fires$P.adj), ]
print(similar_fires[, c("Comparison", "P.adj")])
```

Through our Dunn's test, we have established:

- There are 12 combinations (pairs) of fires that are statistically significantly different in their severity distributions (p<0.05)
- There are 33 combinations (pairs) of fires that are NOT statistically significantly different in their severity distributions (p>0.05).

Now that we have established that there are statistically significant differences between fire severities (Kruskall-Wallis), and we also know from the Dunn's test which combinations/pairs of fires are either statistically significantly different, or NOT statistically significantly different. 

# Forming Fire Clusters
With the above established, let's use Hierarchical Cluster Analysis (HCA), an unsupervised machine learning method that builds a hierarchy of clusters by calculating Euclidean distances in, and presents the results in the form of a dendrogram. 

## HCA Plot
```{r}
fire_medians = aggregate(severity ~ fire, data = ds_fire_og, FUN = median)
dist_matrix = dist(fire_medians$severity)
clusters = hclust(dist_matrix)
plot(clusters, labels = fire_medians$fire, main = "Fire Similarity Groups")
```

In the above HCA plot, we can see the groupings that the unsupervised technique suggests, but it is not as precise as we would like to be going forward. We need to know:

- the exact number of clusters
- what fires should go in which cluster 

Thus, the HCA plot useful going forward, but needs to be refined.

## Total Within-Cluster Sum of Squares Calculation
In order to refine it, we will calculate the "Total Within-Cluster Sum of Squares" for different numbers of clusters (k). In this case, I tested from 1-9 clusters.

```{r}
row.names(fire_medians) = fire_medians$fire
vals = as.matrix(fire_medians$severity)

wss = sapply(1:9, function(k) {
  group = cutree(clusters, k = k)
  sum(sapply(unique(group), function(g) {
    sum((vals[group == g] - mean(vals[group == g]))^2)
  }))
})
```

Once we have our statistic calculated for different cluster numbers, we create a plot with the number of clusters on the x-axis and the aforementioned statistic as the y-axis. By doing this, we are examining the number of clusters that are effectively reducing the variance in the statistic.
```{r}
plot(1:9, wss, type="b", pch = 19, frame = FALSE, 
     xlab="Number of clusters K",
     ylab="Total within-clusters sum of squares",
     main="Elbow Method for Fire Groups")
```
With the plot, we will use the Elbow method, which is essentially how we pick the number of clusters, or groups, that the fires should go in. In the case of this plot, the Elbow method suggests that 3 clusters is the ideal number, as at 3 clusters, we have reduced the statistic significantly, and having 4 or more clusters will lead to unnecessary over-fitting, where we are splitting groups that are already statistically similar. 

Now, we have the groups for the fires as listed below:
```{r}
fire_groups_nums = cutree(clusters, k = 3)
fire_groupings = tibble(
  fire = fire_medians$fire,
  group = fire_groups_nums
)
fire_groupings
```

Now, we will extract the fires in each group and add it to the data.
```{r}
group_1_names = fire_groupings %>%
  filter(group == 1) %>%
  pull(fire)
group_2_names = fire_groupings %>%
  filter(group == 2) %>%
  pull(fire)
group_3_names = fire_groupings %>%
  filter(group == 3) %>%
  pull(fire)

ds_fire = ds_fire_og %>%
  mutate(
    fire_group = case_when(
      fire %in% group_1_names ~ "1",
      fire %in% group_2_names ~ "2",
      fire %in% group_3_names ~ "3"
    ),
    fire_group = factor(fire_group)
  )
```

# Unsupervised Machine Learning
First, we need to do some additional wrangling.

```{r}
# get rid of coordinates and factor all categorical variables
ds_fire = ds_fire %>%
  select(-x, -y) %>%
  mutate(
    fire = factor(fire),
    treated = factor(treated),
    fuel_model = factor(fuel_model),
    vegetation_type = factor(vegetation_type),
    fuel_model_group = factor(fuel_model_group),
    vegetation_type_group = factor(vegetation_type_group)
  ) %>%
  # get rid of severity outliers
  group_by(fire) %>%
  filter(
    severity > quantile(severity, 0.01, na.rm=TRUE),
    severity < quantile(severity, 0.99, na.rm=TRUE)
  ) %>%
  ungroup() %>%
  # classify severity
  mutate(
    severity_class = case_when(
      severity <= -100 ~ "Increased_Vegetation",
      severity >= -100 & severity <= 100 ~ "Low_Severity",
      severity >= 100 & severity <= 270 ~ "Moderate_Severity",
      TRUE ~ "High Severity"
    ),
    severity_class = factor(severity_class)
  ) %>%
  # get rid of missing data
  filter(!is.na(severity_class), 
         !is.na(fire_group),
         !is.na(vpdmax),
         !is.na(elevation_m),
         !is.na(fuel_model_group),
         !is.na(vegetation_type_group)
  )
```


## Non-Metric Multidimensional Scaling (NMDS) - Non-linear:
Non-linear dimensionality reduction technique used to visualize similarity or dissimilarity between complex datasets. Focuses on preserving the rank-order of distances between samples, making it ideal for ecological and high-dimensional data that do not follow linear distributions. Used as a tool for visualization and EDA rather than direct prediction, as it helps identify clusters, patterns, and gradients in data.

The Bray-Curtis distance, which is used to calculate the distance for the NMDS method, requires a numerical matrix. Since the Gower's distance method effectively handles both numerical and categorical data and creates distance between points, it is used on our data prior to feeding the data into the NMDS function.

```{r}
env_data = ds_fire %>%
  select(
    #ppt,
    tmax,
    #vpdmax,
    #canopy_cover,
    #slope_deg,
    #aspect_deg,
    elevation_m,
    fuel_model_group,
    vegetation_type_group) %>%
  mutate(across(where(is.character), as.factor))

dist_matrix = daisy(env_data, metric = "gower")

nmds_result_3d = metaMDS(dist_matrix, k = 3, trymax = 25, autotransform = FALSE)

nmds_matrix_3d = as.matrix(nmds_result_3d$points)

fit = envfit(nmds_matrix_3d, ds_fire[, c("severity_class", "fire_group",
                                      "vpdmax", "elevation_m", 
                                      "fuel_model_group", 
                                      "vegetation_type_group")], 
              perm = 999)

print(fit)
```

Filter Vectors for significance (numerical)
```{r}
sig_vec_idx = which(fit$vectors$pvals < 0.05)
vec_coords_sig = as.data.frame(fit$vectors$arrows[sig_vec_idx, , drop = FALSE])
vec_coords_sig$variable = rownames(vec_coords_sig)
```


Filter Factors for significance(categorical)
```{r}
sig_fact_idx = which(fit$factors$pvals < 0.05)
# This identifies which CATEGORIES (e.g., 'Timber') belong to significant groups
fact_coords_sig = as.data.frame(fit$factors$centroids[sig_fact_idx, , drop = FALSE])
fact_coords_sig$variable = rownames(fact_coords_sig)
```


### Full pipeline to create pairwise panels
Since we have both continuous and categorical factors in our fit variable, we need to extract them before we create the NMDS Visualization.

Step 1: Extract site scores from the 3D NMDS result
```{r}
site_scores = as.data.frame(nmds_result_3d$points)
colnames(site_scores) = c("NMDS1", "NMDS2", "NMDS3")
site_scores$fire_group     = ds_fire$fire_group
site_scores$severity_class = ds_fire$severity_class
```

Step 2: Rerun envfit on the raw points matrix for vectors and factors
```{r}
fit_vec = envfit(
  nmds_result_3d$points,
  ds_fire %>% select(vpdmax, elevation_m),
  perm = 999
)

fit_fac = envfit(
  nmds_result_3d$points,
  ds_fire %>% select(fuel_model_group, vegetation_type_group,
                  severity_class, fire_group),
  perm = 999
)
```

Step 3: Build vec_coords and fact_coords safely
```{r}
vec_coords = as.data.frame(fit_vec$vectors$arrows)
colnames(vec_coords) = c("NMDS1", "NMDS2")
vec_coords$NMDS3    = 0
vec_coords$variable = rownames(vec_coords)

fact_coords <- as.data.frame(fit_fac$factors$centroids)
colnames(fact_coords) = c("NMDS1", "NMDS2")
fact_coords$NMDS3    = 0
fact_coords$variable = rownames(fact_coords)
```

Step 4: Clean up factor labels

Keeping only fuel model centroids -- vegetation type adds too many overlapping labels. Stripping the prefix so "Grass" shows instead of "fuel_model_groupGrass"
```{r}
fact_coords_plot = fact_coords %>%
  filter(grepl("fuel_model_group", variable)) %>%
  mutate(variable = gsub("fuel_model_group", "", variable))
```

Step 5: Shared styling
```{r}
group_colors = c(
  "1" = "#B85042",
  "2" = "#4A7C59",
  "3" = "#4A6FA5"
)

severity_shapes = c(
  "High Severity"        = 17,
  "Increased_Vegetation" = 15,
  "Low_Severity"         = 16,
  "Moderate_Severity"    = 18
)

nmds_theme = theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", size = 14, hjust = 0.5),
    panel.grid.minor   = element_blank(),
    panel.grid.major   = element_line(color = "gray92"),
    panel.border       = element_rect(color = "gray80", fill = NA),
    plot.background    = element_rect(fill = "white", color = NA),
    axis.title         = element_text(size = 11),
    legend.title       = element_text(face = "bold", size = 14),
    legend.text        = element_text(size = 12),
    plot.margin        = unit(c(10, 15, 10, 15), "pt"),
    legend.key.size    = unit(1.2, "lines")  
  )
```

Step 6: Panel builder function
```{r}
make_nmds_panel = function(xvar, yvar) {
  
  pts = site_scores %>% 
    rename(x_axis = all_of(xvar), y_axis = all_of(yvar))
  
  vec = vec_coords %>% 
    rename(x_axis = all_of(xvar), y_axis = all_of(yvar))
  
  fct = fact_coords_plot %>% 
    rename(x_axis = all_of(xvar), y_axis = all_of(yvar))
  
  # Scale arrows relative to the data range so they're always visible
  x_range = diff(range(pts$x_axis, na.rm = TRUE))
  y_range = diff(range(pts$y_axis, na.rm = TRUE))
  arrow_scale = min(x_range, y_range) * 0.35
  
  vec = vec %>%
    mutate(x_end = x_axis * arrow_scale,
           y_end = y_axis * arrow_scale)
  
  ggplot(pts, aes(x = x_axis, y = y_axis,
                  color = fire_group,
                  shape = severity_class)) +
    geom_point(alpha = 0.5, size = 1.5) +
    
    # Environmental vector arrows
    geom_segment(data = vec,
                 aes(x = 0, y = 0, xend = x_end, yend = y_end),
                 inherit.aes = FALSE,
                 arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
                 color = "black", linewidth = 0.9) +
    geom_label(data = vec,
               aes(x = x_end * 1.25, y = y_end * 1.25, label = variable),
               inherit.aes = FALSE,
               size = 5, fontface = "bold", color = "black",
               fill = "white", label.size = 0,
               label.padding = unit(0.15, "lines")) +
    
    # Fuel model centroids with repelled labels
    ggrepel::geom_text_repel(
      data = fct,
      aes(x = x_axis, y = y_axis, label = variable),
      inherit.aes = FALSE,
      color = "red", size = 5, fontface = "bold",
      max.overlaps = 20,
      box.padding = 0.4,
      segment.color = "red", segment.size = 0.3, segment.alpha = 0.5
    ) +
    geom_point(data = fct,
               aes(x = x_axis, y = y_axis),
               inherit.aes = FALSE,
               color = "red", size = 5, shape = 3, stroke = 1.2) +
    
    scale_color_manual(values = group_colors, name = "Fire Group") +
    scale_shape_manual(values = severity_shapes, name = "Severity Class") +
    labs(title = paste(xvar, "vs", yvar), x = xvar, y = yvar) +
    nmds_theme
}
```

Step 7: Build and view each panel individually
```{r}

p1 = make_nmds_panel("NMDS1", "NMDS2")
p2 = make_nmds_panel("NMDS1", "NMDS3")
p3 = make_nmds_panel("NMDS2", "NMDS3")

# View each plot separately
p1
p2
p3
```


Let's look at what significantly drives severity based off NMDS by extracting R² and p-values from the envfit object to create a ranked summary of which environmental variables most strongly align with the ordination space -- and therefore with severity patterns across fires.

Two separate tables are built: one for continuous vectors (tmax, vpdmax, elevation) and one for categorical factors (fuel model, vegetation type, severity class, fire group), since envfit handles them differently

```{r}
# Continuous variables
vector_r2 = as.data.frame(fit$vectors$r)
colnames(vector_r2) = "R2"
vector_r2$p_value = fit$vectors$pvals
vector_r2$variable = rownames(vector_r2)
vector_r2$type = "Continuous"

# Categorical variables
factor_r2 = as.data.frame(fit$factors$r)
colnames(factor_r2) = "R2"
factor_r2$p_value = fit$factors$pvals
factor_r2$variable = rownames(factor_r2)
factor_r2$type = "Categorical"

envfit_summary = bind_rows(vector_r2, factor_r2) %>%
  arrange(desc(R2)) %>%
  mutate(significant = p_value < 0.05)

print(envfit_summary)
```

### Ranked bar chart of R² values by variable
Variables with high R² and significant p-values are the ones most strongly structuring the ordination. These are the best candidates for explaining cross-fire severity patterns in the white paper.

```{r}
envfit_summary = envfit_summary %>%
  mutate(variable_label = case_when(
    variable == "fuel_model_group"      ~ "Fuel Model Type",
    variable == "vegetation_type_group" ~ "Vegetation Type",
    variable == "elevation_m"           ~ "Elevation",
    variable == "vpdmax"                ~ "Atmospheric Dryness (VPD)",
    variable == "severity_class"        ~ "Severity Class",
    variable == "fire_group"            ~ "Fire Group",
    TRUE ~ variable
  ))

ggplot(envfit_summary, 
       aes(x = reorder(variable_label, R2), y = R2)) +
  geom_col(width = 0.65, fill = "#C51616") +
  
  # R² value labels on the end of each bar
  geom_text(aes(label = round(R2, 2)),
            hjust = -0.15,
            fontface = "bold",
            size = 5,
            color = "gray20") +
  
  coord_flip(clip = "off") +
  expand_limits(y = 0.9) +   # gives room for the labels past the bars
  
  labs(
    title    = "What Drives Wildfire Severity Patterns?",
    x        = NULL,         # y-axis label is self explanatory with readable names
    y        = "R² (Proportion of Variance Explained)"
  ) +
  
  theme_minimal(base_size = 15) +
  theme(
    plot.title       = element_text(face = "bold", size = 20, hjust = 0),
    axis.text.y      = element_text(face = "bold", size = 14, color = "gray10"),
    axis.text.x      = element_text(size = 12, color = "gray10"),
    axis.title.x     = element_text(size = 13, color = "gray10",
    panel.grid.major.y = element_blank(),   # removes horizontal gridlines
    panel.grid.major.x = element_line(color = "gray90"),
    legend.position  = "none",
    plot.background  = element_rect(fill = "white", color = NA))
  )
```

Fuel characteristics (fuel model and vegetation type) are the primary structuring forces of fire severity patterns across the landscape. Climate stress (VPDmax) and topography (elevation) play a secondary but still meaningful role.


## Non-Linear Clustering Options: UMAPS, Gower's Distances, HBDSCAN
We use these because linear relationships do not accurately represent the data and the relationship between the variables. 

First, more data wrangling
```{r}
fire_list = ds_fire %>%
  mutate(across(where(is.numeric), ~ifelse(is.infinite(.), NA, .))) %>%
  group_split(fire)
```

Defining the function to run all 3:
```{r}
run_fire_pipeline = function(data){
  fire_name = unique(data$fire)
  feature_cols = c(
    #"ppt",
    "tmax",
    #"vpdmax",
    "canopy_cover",
    #"slope_deg",
    #"aspect_deg",
    "elevation_m",
    "fuel_model_group",
    "vegetation_type_group"
  )
  features = data %>%
    select(all_of(feature_cols)) %>%
    drop_na()
  n = nrow(features)
  
  # Skip fires that are too small
  if(n < 10){
    return(NULL)
  }
  
  features = features %>%
  dplyr::mutate(
    dplyr::across(where(is.character), as.factor)
  )

  # GOWER DISTANCE
  gower_dist = cluster::daisy(features, metric = "gower")
 
  # UMAP (dynamic neighbors)
  neighbors = min(10, n - 1)
  umap_config = umap::umap.defaults
  umap_config$n_neighbors = neighbors
  umap_res = umap::umap(as.matrix(gower_dist), config = umap_config)
  coords = as.data.frame(umap_res$layout)
  colnames(coords) = c("UMAP1","UMAP2")

  # HDBSCAN
  minpts = max(5, floor(n * 0.05))
  hdb = dbscan::hdbscan(coords, minPts = minpts)
  coords$cluster = factor(hdb$cluster)
  coords$fire = fire_name
  cluster_data = data %>%
    slice(1:n) %>%
    mutate(cluster = coords$cluster)
  list(
    coords = coords,
    cluster_data = cluster_data
  )
}
```

Calling the function:
```{r}
results = purrr::map(fire_list, run_fire_pipeline)
results = results[!sapply(results, is.null)]

umap_results = purrr::map_dfr(results, "coords")
cluster_data = purrr::map_dfr(results, "cluster_data")
```

### Visualizing the Results

```{r}
ggplot(umap_results,
       aes(UMAP1, UMAP2, color = cluster)) +
  geom_point(alpha = .7) +
  facet_wrap(~fire) +
  theme_minimal() +
  labs(title = "UMAP Clusters by Fire")

ggplot(cluster_data,
       aes(cluster, severity, fill = cluster)) +
  geom_boxplot() +
  facet_wrap(~fire) +
  theme_minimal() +
  labs(title = "Severity Distribution by Cluster")
```

Summary table:
```{r}
cluster_summary = cluster_data %>%
  group_by(fire, cluster) %>%
  summarise(
    across(
      c(tmax, canopy_cover,
        elevation_m),
      median,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

cluster_summary
```


```{r}
cluster_summary %>%
  pivot_longer(
    -c(fire, cluster),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(cluster, variable, fill = value)) +
    geom_tile() +
    facet_wrap(~fire, scales = "free") +
    scale_fill_viridis_c() +
    theme_minimal()

ggplot(cluster_data,
       aes(cluster, fill = fuel_model_group)) +
  geom_bar(position = "fill") +
  facet_wrap(~fire) +
  theme_minimal() +
  labs(
    y = "Proportion",
    title = "Fuel Model Composition by Cluster"
  )
```

#### Characteristics of each cluster

Take the median of numerical variables, percent of each fuel/veg type by taking mean to make one big table
```{r}
cluster_classifications = cluster_data %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    elevation_m = median(elevation_m, na.rm = TRUE),
    vpdmax = median(vpdmax, na.rm = TRUE),
    canopy_cover = median(canopy_cover, na.rm = TRUE),
    slope_deg = median(slope_deg, na.rm = TRUE),

    # proportions for categorical
    fuel_grass = mean(fuel_model_group == "Grass", na.rm = TRUE),
    fuel_shrub = mean(fuel_model_group == "Shrub", na.rm = TRUE),
    fuel_timberunderstory = mean(fuel_model_group == "Timber-Understory", na.rm = TRUE),
    fuel_timberlitter = mean(fuel_model_group == "Timber Litter", na.rm = TRUE),
    fuel_grassshrub = mean(fuel_model_group == "Grass-Shrub", na.rm = TRUE),
    fuel_nonburn = mean(fuel_model_group == "Non-Burnable", na.rm = TRUE),

    veg_tree = mean(vegetation_type_group == "Tree", na.rm = TRUE),
    veg_shrub = mean(vegetation_type_group == "Shrub", na.rm = TRUE),
    veg_developed = mean(vegetation_type_group == "Developed", na.rm = TRUE),
    veg_herb = mean(vegetation_type_group == "Herb", na.rm = TRUE),
    veg_sparse = mean(vegetation_type_group == "Sparse", na.rm = TRUE),
    veg_water = mean(vegetation_type_group == "Water", na.rm = TRUE),
    veg_ag = mean(vegetation_type_group == "Agriculture", na.rm = TRUE),
    veg_barren = mean(vegetation_type_group == "Barren", na.rm = TRUE),

    severity = median(severity, na.rm = TRUE),

    .groups = "drop"
  )

cluster_classifications
```

- Cluster 0 (33.5 severity): 1292 elevation, 37 vpd, 45 canopy cover, 12 slope, majority timber fuel, majority tree vegetation
- Cluster 1 (-2 severity): 1379 elevation, 39 vpd, 0 canopy cover, 7 slope, majority grass shrub fuel, majority shrub tree vegetation
- Cluster 2 (28.5 severity): 1514 elevation, 36 vpd, 35 canopy cover, 10 slope, majority grass-shrub and timber-litter fuel, majority tree vegetation
- Cluster 3 (56 severity): 1611 elevation, 33 vpd, 45 canopy cover, 10 slope, majority timber fuel, majority tree vegetation
- Cluster 4 (36 severity): 1042 elevation, 42 vpd, 45 canopy cover, 14 slope, majority timber fuel, major majority tree vegetation
- Cluster 5 (31 severity): 967 elevation, 41 vpd, 55 canopy cover, 17 slope, major majority timber fuel, major majority tree vegetation
- Cluster 6 (8 severity): 1024 elevation, 41 vpd, 35 canopy cover, 15 slope, major majority timber understory fuel, major majority tree vegetation
- Cluster 7 (26.5 severity): 1300 elevation, 42 vpd, 35 canopy cover, 15 slope, majority timber understory and grass shrub fuel, majority tree vegetation
- Cluster 8 (27 severity): 1667 elevation, 34 vpd, 45 canopy cover, 19 slope, major majority timber fuel, all tree vegetation
- Cluster 9 (16.5 severity): 1351 elevation, 35 vpd, 45 canopy cover, 12 slope, all timber-litter fuel, all tree vegetation
- Cluster 10 (60 severity): 1715 elevation, 39 vpd, 45 canopy cover, 8 slope, all timber-litter fuel, all tree vegetation
- Cluster 11 (-6.5 severity): 1499 elevation, 43 vpd, 40 canopy cover, 3 slope, all timber-litter fuel, almost all tree vegetation


Cluster analysis revealed that burn severity is structured by distinct environmental–fuel regimes. Higher severity clusters were primarily associated with timber-dominated fuel complexes (including timber litter and understory fuels), moderate canopy cover, and higher elevation forest environments. In contrast, lower severity clusters tended to occur in more open or mixed fuel environments, including grass–shrub and low canopy conditions. These results suggest that fuel continuity and forest structure are more strongly associated with severity patterns than topographic variables such as slope.


Visualize the table:
```{r}
clusters_long = cluster_classifications %>%
  pivot_longer(
    cols = starts_with("fuel_") | starts_with("veg_"),
    names_to = c("fuel_veg", "category"),
    names_sep = "_",
    values_to = "prop"
  )

clusters_long %>%
  filter(cluster == 0) %>%
  ggplot(aes(x = reorder_within(category, -prop, fuel_veg), y = prop)) +
  geom_col() +
  facet_wrap(~fuel_veg, scales = "free_x") +
  scale_x_reordered() +
  theme(axis.text.x = element_text(angle = 90))
  
clusters_long %>%
  filter(fuel_veg == "fuel") %>%
  ggplot(aes(reorder_within(category, -prop, fuel_veg), y = prop, fill = severity)) +
  geom_col() +
  facet_wrap(~cluster) +
  scale_x_reordered() +
  theme(axis.text.x = element_text(angle = 90))

clusters_long %>%
  filter(fuel_veg == "veg") %>%
  ggplot(aes(reorder_within(category, -prop, fuel_veg), y = prop, fill = severity)) +
  geom_col() +
  facet_wrap(~cluster) +
  scale_x_reordered() +
  theme(axis.text.x = element_text(angle = 90))
```


Chi-Squared Test:
```{r}
aov(severity ~ cluster, data = cluster_data)
chisq.test(table(cluster_data$fuel_model_group,
                 cluster_data$cluster))
```



# Hypothesis Testing 

## Treated vs Untreated Severity

h0 = no difference between treated and untreated severity
ha = there is a difference between treated and untreated severity

```{r}
wilcox.test(severity ~ treated, data = ds_fire)
```

p-value > 0.05 -> not enough evidence to reject null

## Severity Among Vegetation Types

h0 = no difference in severity among vegetation types
ha = there is a difference in severity among vegetation types

```{r}
kruskal.test(severity ~ vegetation_type_group, data = ds_fire)

veg_dunn = dunnTest(severity ~ vegetation_type_group,
                     data = ds_fire,
                     method = "holm")

veg_dunn$res[, c("Comparison", "P.adj")] %>%
  arrange(P.adj, desc = T)
```

p-value < 0.05 -> enough evidence to reject null, dunn test to find which vegetation types severity differs for

Dunn test results: significant differences between shrub-tree, herb-tree, agriculture-tree, shrub-water, shrub-sparse, herb-water, agriculture-water

## Severity Among Fuel Types

h0 = no difference in severity among fuel model types
ha = there is a difference in severity among fuel model types

```{r}
kruskal.test(severity ~ fuel_model_group, data = ds_fire)

fuel_dunn = dunnTest(severity ~ fuel_model_group,
                     data = ds_fire,
                     method = "holm")

fuel_dunn$res[, c("Comparison", "P.adj")] %>%
  arrange(P.adj, desc = T)
```

p-value < 0.05 -> enough evidence to reject null, dunn test to find which fuel model types severity differs for

Dunn Test results: all have significant differences but grass-grass-shrub, grass-shrub-shrub, non-burnable-shrub, timber-understory-timber-litter, grass-shrub-non-burnable


## Severity Among Fire Cluster Groups

h0 = our fire cluster groups have no difference in severity
ha = our clustering is meaningful and reflects differences in severity

```{r}
fire_groups_ds = ds_fire %>%
  mutate(
    fire_group = case_when(
      fire %in% group_1_names ~ "1",
      fire %in% group_2_names ~ "2",
      fire %in% group_3_names ~ "3"
    )
  )

kruskal.test(severity ~ fire_group, data = fire_groups_ds)


cluster_dunn = dunnTest(severity ~ fire_group, 
                        method = "holm", 
                        data = fire_groups_ds)

cluster_dunn$res[, c("Comparison", "P.adj")] %>%
  arrange(P.adj, desc = T)
```

p-value < 0.05 -> enough evidence to reject null, dunn test to see which groups are significant

Dunn's Test Results: all groups are significant, so our clustering is meaningful and reflects differences in severity



## Spearman Correlation Test for elevation and vpd
```{r}
cor.test(ds_fire_og$severity, ds_fire_og$elevation_m, method = "spearman")
cor.test(ds_fire_og$severity, ds_fire_og$vpdmax, method = "spearman")

model <- lm(severity ~ elevation_m + vpdmax, data = ds_fire_og)
summary(model)

ds_fire_og %>%
  filter(severity > -2000, severity < 2000) %>%
ggplot(aes(elevation_m, severity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "red") +
  theme_minimal()

ds_fire_og %>%
  filter(severity > -2000, severity < 2000) %>%
ggplot(aes(vpdmax, severity)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", color = "blue") +
  theme_minimal()

```




# Solution Models

## Solution 1: Predicting for severity
Using a random forest model and a binary severity response variable to make it easier to train and interpret
```{r}
# 80/20 train test split
train_idx = sample(1:nrow(ds_fire), 0.8 * nrow(ds_fire))
train_raw = ds_fire[train_idx, ]
test_raw = ds_fire[-train_idx, ]

# making severity binary class from the 4-class variable
test_raw = test_raw %>%
  mutate(
    severity_binary = case_when(
      severity_class == "High Severity" ~ 1,
      severity_class == "Moderate_Severity" ~ 1,
      severity_class == "Low_Severity" ~ 0,
      severity_class == "Increased_Vegetation" ~ 0
    ),
    severity_binary = factor(severity_binary)
  )
train_raw = train_raw %>%
  mutate(
    severity_binary = case_when(
      severity_class == "High Severity" ~ 1,
      severity_class == "Moderate_Severity" ~ 1,
      severity_class == "Low_Severity" ~ 0,
      severity_class == "Increased_Vegetation" ~ 0
    ),
    severity_binary = factor(severity_binary)
  )

# fitting a random forest model
rf_bin = randomForest(
  severity_binary ~ ppt + tmax + vpdmax + elevation_m,
  data = train_raw,
  ntree = 1000,        # more trees = more stable
  mtry = 3,            
  nodesize = 5,
  importance = TRUE
)

# using the test data to determine how good it is
pred = predict(rf_bin, test_raw, type = "prob")[,2]
pred_class = ifelse(pred > 0.5, 1, 0)

mean(pred_class == test_raw$severity_binary)
roc(test_raw$severity_binary, pred)
varImpPlot(rf_bin)
plot(roc(test_raw$severity_binary, pred), auc.polygon = TRUE)
```

AUC measures a model's ability to distinguish between two classes. It represents the probability that a randomly chosen positive instance will be ranked higher by the model than a randomly chosen negative instance. The plot represents the trade-off between Sensitivity (True Positive Rate) and Specificity (1 - False Positive Rate), a curve "hugging" the top-left corner indicates a high AUC


### Generalize to new fires

Data wrangling that happened to the original 10 fire dataset
```{r}
new_fires_og = read_csv("AnalysisDatasets/new_fires_dataset.csv")

new_fires = new_fires_og %>%
  mutate(
    fire = factor(fire),
    treated = factor(treated),
    fuel_model = factor(fuel_model),
    vegetation_type = factor(vegetation_type),
    fuel_model_group = factor(fuel_model_group),
    vegetation_type_group = factor(vegetation_type_group)
  ) %>%
  # get rid of severity outliers
  group_by(fire) %>%
  filter(
    severity > quantile(severity, 0.01, na.rm=TRUE),
    severity < quantile(severity, 0.99, na.rm=TRUE)
  ) %>%
  ungroup() %>%
  # classify severity
  mutate(
    severity_class = case_when(
      severity <= -100 ~ "Increased_Vegetation",
      severity >= -100 & severity <= 100 ~ "Low_Severity",
      severity >= 100 & severity <= 270 ~ "Moderate_Severity",
      TRUE ~ "High Severity"
    ),
    severity_class = factor(severity_class)
  ) %>%
  # get rid of missing data
  filter(!is.na(severity_class), 
         !is.na(vpdmax),
         !is.na(elevation_m),
         !is.na(fuel_model_group),
         !is.na(vegetation_type_group)
  )

new_fires = new_fires %>%
  mutate(
    severity_binary = case_when(
      severity_class == "High Severity" ~ 1,
      severity_class == "Moderate_Severity" ~ 1,
      severity_class == "Low_Severity" ~ 0,
      severity_class == "Increased_Vegetation" ~ 0
    ),
    severity_binary = factor(severity_binary)
  )
```

Making predictions with the new fires
```{r}
pred_new_fires = predict(rf_bin, new_fires, type = "prob")[,2]
pred_class_new_fires = ifelse(pred_new_fires > 0.5, 1, 0)

new_fires$pred_prob = pred_new_fires
new_fires$pred_class = pred_class_new_fires

# calculating some stats to use in subsequent plots
new_fires = new_fires %>%
  mutate(
    mismatch = ifelse(severity_binary == 1 & treated == 0, 1, 0),
    case = case_when(
      severity_binary == 1 & treated == 1 ~ "High severity + treated",
      severity_binary == 1 & treated == 0 ~ "High severity + not treated",
      severity_binary == 0 & treated == 1 ~ "Low severity + treated",
      TRUE ~ "Low severity + not treated"
    ),
    correct = ifelse(pred_class == severity_binary, 1, 0),
    model_helpful = ifelse(pred_class == 1 & severity_binary == 1 & treated == 0,1, 0),
    model_worse = ifelse(pred_class == 0 & severity_binary == 1 & treated == 1, 1, 0),
    case_type = case_when(
      model_worse == 1 ~ "Model worse",
      model_helpful == 1 ~ "Model helpful",
      TRUE ~ "Other"
    )
  )
```


Visualizing results with the new fires

Severity risk vs treatment locations
```{r}
McCash_subset = new_fires %>%
  filter(fire == "McCash_2021") 
Park_subset = new_fires %>%
  filter(fire == "Park_2024")

ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "McCash Fire (2021)")


ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  geom_point(
    data = subset(McCash_subset, treated == 1),
    color = "cyan",
    size = 0.7,
    alpha = 0.7
  ) +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "Park Fire (2024)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = pred_prob), size = 0.5) +
  scale_color_viridis_c(name = "Predicted Risk") +
  geom_point(
    data = subset(Park_subset, treated == 1),
    color = "cyan",
    size = 0.7,
    alpha = 0.7
  ) +
  theme_minimal() +
  labs(title = "Predicted Wildfire Severity Risk vs Treatment Locations",
       subtitle = "Park Fire (2024)")
```


Predicted class
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(pred_class)), size = 0.5) +
  theme_minimal() +
  labs(color = "Predicted Class",
       subtitle = "McCash Fire (2021)")



ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(pred_class)), size = 0.5) +
  theme_minimal() +
  labs(color = "Predicted Class",
       subtitle = "Park Fire (2024)")
```


High-risk areas without treatment
```{r}

ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(mismatch)), size = 0.5) +
  scale_color_manual(values = c("0" = "grey80", "1" = "red"),
                     labels = c("OK", "High Risk Untreated")) +
  theme_minimal() +
  labs(title = "High-Risk Areas Without Treatment",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(mismatch)), size = 0.5) +
  scale_color_manual(values = c("0" = "grey80", "1" = "red"),
                     labels = c("OK", "High Risk Untreated")) +
  theme_minimal() +
  labs(title = "High-Risk Areas Without Treatment",
       subtitle = "Park Fire (2024)")
```


Actual wildfire severity with treatment locations
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(severity_binary)), size = 0.5) +
  geom_point(
    data = subset(McCash_subset, treated == 1),
    color = "cyan",
    size = 0.6,
    alpha = 0.6
  ) +
  scale_color_manual(values = c("grey80", "red"),
                     labels = c("Low severity", "High severity")) +
  theme_minimal() +
  labs(title = "Actual Wildfire Severity with Treatment Locations",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(severity_binary)), size = 0.5) +
  geom_point(
    data = subset(Park_subset, treated == 1),
    color = "cyan",
    size = 0.6,
    alpha = 0.6
  ) +
  scale_color_manual(values = c("grey80", "red"),
                     labels = c("Low severity", "High severity")) +
  theme_minimal() +
  labs(title = "Actual Wildfire Severity with Treatment Locations",
       subtitle = "Park Fire (2024)")
```


model prediction accuracy across landscape
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(correct)), size = 0.5) +
  scale_color_manual(values = c("red", "green"),
                     labels = c("Incorrect", "Correct")) +
  theme_minimal() +
  labs(title = "Model Prediction Accuracy Across Landscape",
       subtitle = "McCash Fire (2021)")


ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = factor(correct)), size = 0.5) +
  scale_color_manual(values = c("red", "green"),
                     labels = c("Incorrect", "Correct")) +
  theme_minimal() +
  labs(title = "Model Prediction Accuracy Across Landscape",
       subtitle = "Park Fire (2024)")
```


model effectiveness
```{r}
ggplot(McCash_subset, aes(x = x, y = y)) +
  geom_point(aes(color = case_type), size = 0.5) +
  scale_color_manual(
    values = c(
      "Model helpful" = "green",
      "Model worse" = "red",
      "Other" = "grey85"
    )
  ) +
  theme_minimal() +
  labs(
    title = "Model Effectiveness for the McCash Fire (2021)",
    subtitle = "Red = high severity + treated areas missed by model",
    color = "Model Outcome"
  )



ggplot(Park_subset, aes(x = x, y = y)) +
  geom_point(aes(color = case_type), size = 0.5) +
  scale_color_manual(
    values = c(
      "Model helpful" = "green",
      "Model worse" = "red",
      "Other" = "grey85"
    )
  ) +
  theme_minimal() +
  labs(
    title = "Model Effectiveness for the Park Fire (2024)",
    subtitle = "Red = high severity + treated areas missed by model",
    color = "Model Outcome"
  )
```


determining the efficiency of treatment for model and actual
```{r}

compute_treatment_efficiency <- function(df, fire_name) {

  n_treated <- sum(df$treated == 1)

  real_hit_rate <- df %>%
    filter(treated == 1) %>%
    summarise(hit = mean(severity_binary == 1)) %>%
    pull(hit)

  model_treated <- df %>%
    arrange(desc(pred_prob)) %>%
    slice(1:n_treated)

  model_hit_rate <- mean(model_treated$severity_binary == 1)

  data.frame(
    fire = fire_name,
    method = c("Actual treated area", "Model-based treated area"),
    hits = c(real_hit_rate, model_hit_rate)
  )
}

results <- bind_rows(
  compute_treatment_efficiency(McCash_subset, "McCash (2021)"),
  compute_treatment_efficiency(Park_subset, "Park (2024)")
)


ggplot(results, aes(x = method, y = hits, fill = method)) +
  geom_bar(stat = "identity") +
  facet_wrap(~fire) +
  theme_bw() +
  labs(
    title = "Efficiency of Treatment Placement Across Fires",
    y = "High-severity pixels captured",
    x = NULL
  ) +
  scale_y_continuous(labels=scales::percent_format(), limits = c(0, 1)) +
  theme(legend.position = "none")
```




## Solution 2: Clustering/Classification Model

post-hoc classification of unsupervised structure, testing whether clusters correspond to predictable environmental regimes - Can we reconstruct these groupings from raw variables?

testing stability and separability of clusters in feature space
```{r}
cluster_data = na.omit(cluster_data)
train_idx = sample(1:nrow(cluster_data), 0.8 * nrow(cluster_data))
train = cluster_data[train_idx, ]
test  = cluster_data[-train_idx, ]

predictors = c(
  "elevation_m",
  "canopy_cover",
  "slope_deg",
  "vpdmax",
  "tmax",
  "fuel_model_group",
  "vegetation_type_group",
  "ppt",
  "aspect_deg"
)
```


```{r}
model1 = randomForest(cluster ~ ., data = train[, c(predictors, "cluster")])
pred1 = predict(model1, test)
mean(pred1 == test$cluster)
cm1 = table(Predicted = pred1, Actual = test$cluster)
cm1
diag(cm1) / colSums(cm1)

varImpPlot(model1)
```



```{r}
model2 = randomForest(
  cluster ~ .,
  data = train[, c(predictors, "cluster")],
  strata = train$cluster,
  sampsize = rep(min(table(train$cluster)), length(unique(train$cluster)))
)
pred2 = predict(model2, test)
mean(pred2 == test$cluster)

cm2 = table(Predicted = pred2, Actual = test$cluster)
cm2
diag(cm2) / colSums(cm2)
```


```{r}
train_control = trainControl(method = "cv", number = 5)

model3 = train(
  cluster ~ .,
  data = train[, c(predictors, "cluster")],
  method = "rf",
  trControl = train_control
)

pred3 = predict(model3, test)
mean(pred3 == test$cluster)
cm3 = table(Predicted = pred3, Actual = test$cluster)
cm3
diag(cm3) / colSums(cm3)

plot(varImp(model3))
```


Cross-validated classification accuracy was approximately 67%, indicating moderate but meaningful separability of unsupervised fire regime clusters based on environmental predictors.


### White Paper Solution 2 Figure
Make a visual table of cluster classifications with characteristics and recommended treatments for each cluster
```{r}
# turn it into a table then dataframe to make it easier to work with
cluster_t = as.data.frame(t(cluster_classifications), stringsAsFactors = FALSE)

# make the rownames under one 'variable' column
cluster_t = tibble::rownames_to_column(cluster_t, var = "variable")
# take the cluster section of variable out, it's already implied
cluster_t = cluster_t[cluster_t$variable != "cluster", ]
# take the V out of the cluster column names
colnames(cluster_t) = gsub("^V", "", colnames(cluster_t))

# arrange by severity
cluster_t = cluster_t %>%
  arrange(desc(variable == "severity"))

# turn variables into numberic and round to hundredth
cluster_t[,-1] = lapply(cluster_t[,-1], as.numeric)
cluster_t[,-1] = lapply(cluster_t[,-1], function(x) round(as.numeric(x), 2))

# rename all of the 'variable' rows so the plot is easier to read
cluster_t = cluster_t %>%
  mutate(variable = recode(variable,
    severity = "Severity",
    n = "Number of Points",
    vpdmax = "VPD",
    elevation_m = "Elevation",
    canopy_cover = "Canopy Cover",
    slope_deg = "Slope",
    aspect_deg = "Aspect",
    fuel_grass = "Fuel: Grass",
    fuel_shrub = "Fuel: Shrub",
    fuel_timberunderstory = "Fuel: Timer-Understory",
    fuel_timberlitter = "Fuel: Timber-Litter",
    fuel_grassshrub = "Fuel: Grass-Shrub",
    fuel_nonburn = "Fuel: Non-Burnable",
    veg_shrub = "Veg: Shrub",
    veg_tree = "Veg: Tree",
    veg_developed = "Veg: Developed",
    veg_herb = "Veg: Herb",
    veg_sparse = "Veg: Sparse",
    veg_water = "Veg: Water",
    veg_ag = "Veg: Agriculture",
    veg_barren = "Veg: Barren"
  ))

# pivot longer
df = cluster_t
df[,-1] <- lapply(df[,-1], as.numeric)
df_long = df %>%
  pivot_longer(
    cols = -variable,
    names_to = "cluster",
    values_to = "value"
  )

# scale the values of each row relative to each other for the coloring in the plot
df_long = df_long %>%
  group_by(variable) %>%
  mutate(
    scaled = (value - min(value, na.rm = TRUE)) /
             (max(value, na.rm = TRUE) - min(value, na.rm = TRUE))
  ) %>%
  ungroup()

# get rid of unnecessary variables
df_long = df_long %>%
  filter(!variable %in% c("Veg: Barren", "Number of Points", "Severity"))

# set the order that variable and cluster should appear in
df_long$variable = factor(
  df_long$variable,
  levels = c(
    "VPD",
    "Elevation",
    "Slope",
    "Aspect",
    "Canopy Cover",
    "Fuel: Grass",
    "Fuel: Shrub",
    "Fuel: Timer-Understory",
    "Fuel: Timber-Litter",
    "Fuel: Grass-Shrub",
    "Fuel: Non-Burnable",
    "Veg: Shrub",
    "Veg: Tree",
    "Veg: Developed",
    "Veg: Herb",
    "Veg: Sparse",
    "Veg: Water",
    "Veg: Agriculture"
  )
)
df_long$variable = factor(df_long$variable, levels = rev(levels(df_long$variable)))

# set the order of the clusters so they are grouped by similar environment
cluster_order = c(
  "1","5","2",      # shrub/grass
  "3","4",          # transitional
  "6","7","9","10", # forest
  "8","11"          # low fuel / edge
)

df_long$cluster = factor(df_long$cluster, levels = cluster_order)


# final visualization of the heatmap part of the figure
heatmap_plot = ggplot(df_long, aes(x = cluster, y = variable, fill = scaled)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "firebrick1") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  theme_bw() +
  labs(
    x = NULL,
    y = "Input Variable",
    fill = "Relative\nIntensity"
  ) +
  scale_x_discrete(position = "top") +
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.title.x.top = element_blank(),
    axis.text.x = element_blank(),
    axis.text.x.top = element_blank(),
    axis.ticks.x = element_blank(),
    axis.ticks.x.top = element_blank()
  )


# bottom part of the figure, treatments for each cluster
treatments = data.frame(
  cluster = factor(1:11),
  treatment = c(
    "Rx Fire + Grazing",
    "Rx Fire + Fuel Breaks",
    "Thin + Rx Fire",
    "Rx Fire + Patch Thin",
    "Rx Fire + Fuel Breaks",
    "Thin + Rx Fire",
    "Thin + Rx Fire",
    "Defensible Space + Fuel Breaks",
    "Intensive Thin + Fuel Breaks",
    "Thin + Managed Fire",
    "Edge Thin + Defensible Space"
  )
)

treatments$cluster = factor(treatments$cluster, levels = levels(df_long$cluster))
treatments$treatment = str_wrap(treatments$treatment, width = 7)

# making the bottom plot with all the treatments
p_treat = ggplot(treatments, aes(x = cluster, y = 1, label = treatment)) +
  geom_vline(xintercept = seq(1.5, 10.5, by = 1), color = "grey80", linewidth = 0.4) +
  geom_text(
    aes(label = treatment),
    size = 2.3,
    hjust = 0,
    nudge_x = -0.35
  ) +
  theme_void() +
  theme(
    plot.margin = ggplot2::margin(1, 1, 1, 1),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 9
    ),
    axis.title.y = element_text(
  angle = 0,
  vjust = 0.5,
  hjust = 1,                  # anchor closer to plot
  margin = ggplot2::margin(r = -100),   # pull it inward (negative = closer)
  size = 10
)
  ) +
  labs(y = "Recommended Treatment")


# creating the labels for the cluster groupings to go above the heatmap
group_labels = tibble(
  group = c("Shrub / Grass", "Transitional", "Forest", "Low Fuel / Edge"),
  xmin = c(0.5, 3.5, 5.5, 9.5),
  xmax = c(3.5, 5.5, 9.5, 11.5),
  xmid = (xmin + xmax) / 2
)

# plot for the cluster groupings
group_bar = ggplot(group_labels) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 1),
            fill = "grey90", color = "white") +
  geom_text(aes(x = xmid, y = 0.5, label = group),
            size = 2.5, fontface = "bold") +
  scale_x_continuous(
    limits = c(0.5, length(cluster_order) + 0.5),
    breaks = seq_along(cluster_order),
    labels = cluster_order,
    expand = c(0, 0),
    position = "top"
  ) +
  labs(x = "Cluster") +
  theme_void() +
  theme(
    axis.text.x.top = element_text(size = 9),
    axis.title.x.top = element_text(size = 10, face = "bold", margin = ggplot2::margin(b = 4)),
    plot.margin = ggplot2::margin(0, 5, 0, 5)
  )

# putting all plots together
final_plot = group_bar / heatmap_plot / p_treat +
  plot_layout(
    heights = c(0.55, 4, 1),
    guides = "collect"
  ) +
  plot_annotation(
    title = "Fire Regime Environmental Profiles and Associated Treatments"
  ) &
  theme(
    legend.position = "top",
    legend.box = "horizontal",
    legend.key.height = unit(0.1, "cm"),
    legend.key.width = unit(1, "cm"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8, margin = ggplot2::margin(r = 10)),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

final_plot
```




# Recommendations figure

```{r}
# setting up structure and text of the flowchart
nodes = data.frame(
  x = c(1,1,1,1,1,1,1),
  y = c(7,6,5,4,3,2,1),
  label = c(
    "Severity, Treatment, and Environment Data",
    "Random Forest Risk Model (Solution 1)",
    "High-Risk Areas Identified",
    "Regime Classification Model (Solution 2)",
    "Regime-Specific Treatment Selection",
    "Targeted Treatment Implementation",
    "Reduced High-Severity Fire Outcomes"
  )
)

# making the last one bold
nodes$label[7] = expression(bold("Reduced High-Severity Fire Outcomes"))

# plotting it as a segment graph
ggplot(nodes, aes(x = x, y = y)) +
  geom_text(aes(label = label), size = 4, lineheight = 0.9) +
  geom_segment(aes(x = 1, xend = 1, y = 6.8, yend = 6.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 5.8, yend = 5.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 4.8, yend = 4.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 3.8, yend = 3.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 2.8, yend = 2.2), arrow = arrow(length = unit(0.2, "cm"))) +
  geom_segment(aes(x = 1, xend = 1, y = 1.8, yend = 1.2), arrow = arrow(length = unit(0.2, "cm"))) +
  theme_void() +
  labs(
    title = expression(bold("Integrated Decision Framework for Wildfire Treatment Allocation and Design"))
  ) +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
```










